---
output:
  html_document: default
  pdf_document: default
---
### Integration with public data

---
  title: "Finding gene modules that correlate with neurite growth"
output:
  html_document: default
github_document: default
fig_width: 12
fig_height: 4 
---
  
```{r include = FALSE}
### Load data and packages
knitr::opts_chunk$set(fig.width=24, fig.height=8) 

```

In this script I integrate the mouse neuron data with publically available datasets and compare PCA and WGCNA for gene module discovery.

Load the required R libraries:
```{r, include = TRUE, message = FALSE, warning = FALSE, cache = FALSE}
library(readxl)
library(dplyr)
library(Seurat)
library(myUtils)
library(EnsDb.Mmusculus.v75)
library(lubridate)
library(ComplexHeatmap)
library(R.matlab)
library(WGCNA)
library(biomaRt)
library(org.Mm.eg.db)
library(gridExtra)
library(grid)
```

Here I 1.) load the data 2.) change the gene identifiers from ensemble ids to symbols, where this is possible (otherwise I leave the ensembl id in place) 3.) load the metadata 4.) I put the data and metadata into a Seurat object. 5.) I also load mitochondrial mouse genes from a database for later use.
```{r, include = TRUE, message = FALSE, warning = FALSE, cache = FALSE}
directory = '/home/jovyan/axonGrowth/'
setwd(directory)
axonGrowth_counts = read.csv('counts.csv', row.names = 1)
rownames(axonGrowth_counts) = unlist(lapply(1:length(rownames(axonGrowth_counts)), function(x) strsplit(rownames(axonGrowth_counts)[x], split = '\\.')[[1]][1]))
symbols = mapIdsMouse(rownames(axonGrowth_counts), 'ENSEMBL', 'SYMBOL')
mitogenes <- genes(EnsDb.Mmusculus.v75, filter = ~ seq_name == "MT")$gene_id
percent.mt = colSums(axonGrowth_counts[rownames(axonGrowth_counts) %in% mitogenes,])/colSums(axonGrowth_counts)
rownames(axonGrowth_counts) = symbols
#sum(unlist(lapply(1:dim(axonGrowth_counts)[1], function(x) substring(rownames(axonGrowth_counts)[x], 1,3))) != 'ENS')/dim(axonGrowth_counts)[1]
metadata = read_xlsx('Summary-neurons_microcoverslips_RNAseq.xlsx', col_types = c('text', 'text', 'date', 'date', 'date', 'date', 'date', rep('text', 7)), skip = 1)
metadata = metadata[10:105,]
axonGrowth = CreateSeuratObject(axonGrowth_counts, project = 'axonGrowth', min.cells = 0, min.features = 0)
axonGrowth$TotalNeuriteLength = as.numeric(metadata$`Total Neurite Length (um)`)
axonGrowth$LongestNeuriteLength = as.numeric(metadata$`Longest Neurite Length (um)`)
axonGrowth$TimeSincePlating = unlist(lapply(1:dim(metadata)[1], function(x) interval(ymd_hms(metadata$`Time of plating (mm/dd/yyyy hh:mm)`[x]), ymd_hms(metadata$`Cell collection time (mm/dd/yyyy hh:mm)`[x]))/dhours(1)))
axonGrowth$TimeSincePlating = unlist(lapply(1:dim(metadata)[1], function(x) interval(ymd_hms(metadata$`Time of plating (mm/dd/yyyy hh:mm)`[x]), ymd_hms(metadata$`Cell collection time (mm/dd/yyyy hh:mm)`[x]))/dhours(1)))
axonGrowth$TimeInAIRMEM = unlist(lapply(1:dim(metadata)[1], function(x) interval(ymd_hms(metadata$`AIR-MEM change (mm/dd/yyyy hh:mm)`[x]), ymd_hms(metadata$`Cell collection time (mm/dd/yyyy hh:mm)`[x]))/dhours(1)))
```

The QC plots using number of detected genes, number of counts and percent of counts coming from mitochondrial genes (as a proxy for stress), show a couple of outlier cells, which I remove in the last line:

```{r, include = TRUE, message = FALSE, warning = FALSE, cache = FALSE, fig.height = 8, fig.width = 16}

axonGrowth[["percent.mt"]] <- percent.mt
VlnPlot(axonGrowth, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
plot1 <- FeatureScatter(axonGrowth, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(axonGrowth, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
CombinePlots(plots = list(plot1, plot2))
axonGrowth <- subset(axonGrowth, subset = nFeature_RNA > 10000 & nFeature_RNA < 20000 & percent.mt < 2.5 & nCount_RNA > 10^5 & nCount_RNA < 10^6)
VlnPlot(axonGrowth, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

```

The following normalizes, scales and select 1000 particularly variable genes. Doing the downstream analysis with only the 1000 most variable genes seems like a waste of information. However, given we only have 85 high quality cells available for this analysis at the moment, it will give us the most robust clustering results. 

```{r, include = TRUE, message = FALSE, warning = FALSE, cache = FALSE, fig.height = 8, fig.width = 16}
axonGrowth <- NormalizeData(axonGrowth, normalization.method = "LogNormalize", scale.factor = 10000)
axonGrowth <- FindVariableFeatures(axonGrowth, selection.method = "vst", nfeatures = 1000)
top25 <- head(VariableFeatures(axonGrowth), 25)
plot1 <- VariableFeaturePlot(axonGrowth)
plot2 <- LabelPoints(plot = plot1, points = top25, repel = TRUE)
CombinePlots(plots = list(plot1, plot2))
all.genes <- rownames(axonGrowth)
axonGrowth <- ScaleData(axonGrowth, features = all.genes)
```

These are the results of the PCA analysis:

```{r, include = TRUE, message = FALSE, warning = FALSE, cache = FALSE, fig.height = 8, fig.width = 16}
axonGrowth <- RunPCA(axonGrowth, features = VariableFeatures(object = axonGrowth))
print(axonGrowth[["pca"]], dims = 1:5, nfeatures = 5)
VizDimLoadings(axonGrowth, dims = 1:2, reduction = "pca")
DimPlot(axonGrowth, reduction = "pca")
DimHeatmap(axonGrowth, dims = 1, cells = round(dim(axonGrowth)[2])/2, balanced = TRUE)
DimHeatmap(axonGrowth, dims = 1:15, cells = round(dim(axonGrowth)[2])/2, balanced = TRUE)
```

Based on the JackStraw procedure I select 12 PCs for further analysis:

```{r, include = TRUE, message = FALSE, warning = FALSE, cache = FALSE, fig.height = 8, fig.width = 16}
axonGrowth <- JackStraw(axonGrowth, num.replicate = 100)
axonGrowth <- ScoreJackStraw(axonGrowth, dims = 1:20)
JackStrawPlot(axonGrowth, dims = 1:20)
ElbowPlot(axonGrowth, ndims = 40)
n_components = 12
```

Let's load the data from https://science.sciencemag.org/content/360/6385/176 as a reference:

```{r, include = TRUE, message = FALSE, warning = FALSE, cache = FALSE, fig.height = 8, fig.width = 16}
reference = readMat('/home/jovyan/data/developingMouse/GSM3017261_150000_CNS_nuclei.mat')
```

We focus on hippocampal neurons from the mouse brain:

```{r, include = TRUE, message = FALSE, warning = FALSE, cache = FALSE, fig.height = 8, fig.width = 16}
region = unlist(lapply(reference$cluster.assignment, function(x) strsplit(x, split = ' ')[[1]][2]))

subset = which(region == 'HIPP')

reference$DGE = reference$DGE[subset,]
reference$barcodes = reference$barcodes[subset]
reference$cluster.assignment = reference$cluster.assignment[subset]
reference$sample.type = reference$sample.type[subset]
reference$genes = trimws(reference$genes)
subset_genes = which(reference$genes %in% rownames(axonGrowth))
reference$genes = reference$genes[subset_genes]
reference$DGE = reference$DGE[,subset_genes]

reference_matrix = as.matrix(reference$DGE)
reference_matrix = t(reference_matrix)
rownames(reference_matrix) = reference$genes
colnames(reference_matrix) = reference$barcodes

reference$DGE = NULL

Reference = CreateSeuratObject(reference_matrix)

Reference <- NormalizeData(Reference, verbose = FALSE)
Reference <- FindVariableFeatures(Reference, selection.method = "vst", nfeatures = 2000, verbose = FALSE)

```

Now we integrate the two datasets using the method provided by Seurat (see https://www.sciencedirect.com/science/article/pii/S0092867419305598?via%3Dihub):

```{r, include = TRUE, message = FALSE, warning = FALSE, cache = FALSE, fig.height = 8, fig.width = 16}
axonGrowth.anchors <- FindIntegrationAnchors(object.list = list(Reference, axonGrowth), dims = 1:30, verbose = TRUE, k.filter = 77)
axonGrowth.integrated <- IntegrateData(anchorset = axonGrowth.anchors, dims = 1:30)
```

Let's vizualize differentiation trajectories with the markers suggested in the reference paper and see how our data relates to the reference data on a UMAP plot:

```{r, include = TRUE, message = FALSE, warning = FALSE, cache = FALSE, fig.height = 8, fig.width = 16}


# Run the standard workflow for visualization and clustering
axonGrowth.integrated <- ScaleData(axonGrowth.integrated, verbose = FALSE)
axonGrowth.integrated <- RunPCA(axonGrowth.integrated, npcs = 30, verbose = FALSE)
axonGrowth.integrated <- RunUMAP(axonGrowth.integrated, reduction = "pca", dims = 1:5)
axonGrowth.integrated$orig.ident[axonGrowth.integrated$orig.ident == 'SeuratProject'] = 'Reference'
axonGrowth.integrated$orig.ident[axonGrowth.integrated$orig.ident == 'axonGrowth'] = 'Our data'

DimPlot(axonGrowth.integrated, reduction = "umap", group.by = "orig.ident")
FeaturePlot(axonGrowth.integrated, features = c('Mki67', 'Prox1', 'Dock10', 'Meis2', 'Spock1'))
FeaturePlot(axonGrowth.integrated, features = c('Mki67', 'Prox1', 'Dock10', 'Meis2', 'Spock1'), cells = colnames(axonGrowth.integrated)[axonGrowth.integrated$orig.ident == 'Our data'])


```

Now we find correlated gene modules with WGCNA, see https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/index.html and see to what extent their eigengene correlates with total neurite length and longest neurite length:

```{r, include = TRUE, message = FALSE, warning = FALSE, cache = FALSE, fig.height = 8, fig.width = 16}
options(stringsAsFactors = FALSE);
enableWGCNAThreads()
# Choose a set of soft-thresholding powers
powers = c(c(1:10), seq(from = 12, to=20, by=2))
# Call the network topology analysis function
sft = pickSoftThreshold(t(axonGrowth.integrated@assays$integrated@scale.data), powerVector = powers, verbose = 5)
# Plot the results:
sizeGrWindow(9, 5)
par(mfrow = c(1,2));
cex1 = 0.9;
# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
main = paste("Scale independence"));
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
labels=powers,cex=cex1,col="red");
# this line corresponds to using an R^2 cut-off of h
abline(h=0.90,col="red")
# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,5],
xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
main = paste("Mean connectivity"))
text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")
softPower = 5;
adjacency = adjacency(t(axonGrowth.integrated@assays$integrated@scale.data), power = softPower)
TOM = TOMsimilarity(adjacency);
dissTOM = 1-TOM
geneTree = hclust(as.dist(dissTOM), method = "average");
# Plot the resulting clustering tree (dendrogram)
sizeGrWindow(12,9)
plot(geneTree, xlab="", sub="", main = "Gene clustering on TOM-based dissimilarity",
labels = FALSE, hang = 0.04);
# We like large modules, so we set the minimum module size relatively high:
minModuleSize = 30;
# Module identification using dynamic tree cut:
dynamicMods = cutreeDynamic(dendro = geneTree, distM = dissTOM,
deepSplit = 2, pamRespectsDendro = FALSE,
minClusterSize = minModuleSize);
table(dynamicMods)
# Convert numeric lables into colors
dynamicColors = labels2colors(dynamicMods)
table(dynamicColors)
# Plot the dendrogram and colors underneath
sizeGrWindow(8,6)
plotDendroAndColors(geneTree, dynamicColors, "Dynamic Tree Cut",
dendroLabels = FALSE, hang = 0.03,
addGuide = TRUE, guideHang = 0.05,
main = "Gene dendrogram and module colors")

datTraits = data.frame(TotalNeuriteLength = axonGrowth.integrated$TotalNeuriteLength,
                       LongestNeuriteLength = axonGrowth.integrated$LongestNeuriteLength)

MEs0 = moduleEigengenes(t(axonGrowth.integrated@assays$integrated@scale.data), dynamicColors)$eigengenes
MEs = orderMEs(MEs0)
moduleTraitCor = cor(MEs, datTraits, use = "p");
nSamples = dim(axonGrowth.integrated@assays$integrated@scale.data)[2]
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nSamples);

sizeGrWindow(10,6)
# Will display correlations and their p-values
textMatrix = paste(signif(moduleTraitCor, 2), "\n(",
signif(moduleTraitPvalue, 1), ")", sep = "");
dim(textMatrix) = dim(moduleTraitCor)
par(mar = c(6, 8.5, 3, 3));
# Display the correlation values within a heatmap plot
labeledHeatmap(Matrix = moduleTraitCor,
xLabels = names(datTraits),
yLabels = names(MEs),
ySymbols = names(MEs),
colorLabels = FALSE,
colors = greenWhiteRed(50),
textMatrix = textMatrix,
setStdMargins = FALSE,
cex.text = 0.5,
zlim = c(-1,1),
main = paste("Module-trait relationships"))
```

Finally let's see how the gene modules relate to the UMAP plot:

```{r, include = TRUE, message = FALSE, warning = FALSE, cache = FALSE, fig.height = 8, fig.width = 16}

axonGrowth.integrated$MEturquoise = MEs$MEturquoise
axonGrowth.integrated$MEblue = MEs$MEblue
axonGrowth.integrated$MEbrown = MEs$MEbrown
axonGrowth.integrated$MEgrey = MEs$MEgrey

FeaturePlot(axonGrowth.integrated, features = c('MEturquoise', 'MEblue', 'MEbrown', 'MEgrey'))
```

And what go functions they are enriched for:

```{r, include = TRUE, message = FALSE, warning = FALSE, cache = FALSE, fig.height = 8, fig.width = 16}
library(topGO)

for (module in c('turquoise', 'blue', 'brown', 'grey')){
  print(module)
  all_genes = rep(0,dim(axonGrowth.integrated@assays$integrated@scale.data)[1])
names(all_genes) = rownames(axonGrowth.integrated@assays$integrated@scale.data)
all_genes[dynamicColors == module] = 1
GOdata <- new("topGOdata", ontology = "BP", allGenes = all_genes, geneSel = function(p) p < 
    0.05, description = "Test", annot = annFUN.org, mapping = "org.Mm.eg.db", 
    ID = "symbol")

resultFisher <- runTest(GOdata, algorithm = "classic", statistic = "fisher")
tab = GenTable(GOdata, classicFisher = resultFisher, topNodes = 10)
d = as.data.frame(as.matrix(tab[,c(1,2,6)]))
p = grid.table(d)
print(p)
print(d)
}

```

The 'blue module' is moderately correlated with Neurite Length and enriched for known GO processes involved in neurite growth. The genes contained in this module are printed out below:

```{r, include = TRUE, message = FALSE, warning = FALSE, cache = FALSE, fig.height = 8, fig.width = 16}

print(names(all_genes)[dynamicColors == 'blue'])

```

However, the below plot that shows Longest Neurite Length as a function of this modules eigengene (PC1), shows no clear association between the two:

```{r, include = TRUE, message = FALSE, warning = FALSE, cache = FALSE, fig.height = 8, fig.width = 16}

plot(axonGrowth.integrated$MEblue, axonGrowth.integrated$TotalNeuriteLength, pch = '.', cex = 10, xlab = 'Blue Module PCA1-score', ylab = 'LongestNeuriteLength')

```

